 

### 2020-09-21 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF). 
There is also increasing severity and frequency of heat waves (Perkins et al. 2012).
This is expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will have a critical influence on the future of trajectory of climate change [@bonan_ecological_2016, Bonan 2008].
With increases in temperature, temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018, Huang et al. 2019]. 
This, alongside climate change-driven disturbances such as drought, may contribute to increasing tree mortality in forests around the world [@mcdowellMechanismsPlantSurvival2008, @allenUnderestimationGlobalVulnerability2015, Brienen et al. 2015].

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**
Globally, we're seeing larger trees suffering more during drought [@bennettLargerTreesSuffer2015, @stovallTreeHeightExplains2019] 
Increasing mortality of large trees, sometimes coupled with reduced tree growth rates, is driving a global trend towards more disturbed forests with younger trees (McDowell et al.2020). 
Additionally, decrease in canopy cover or increase in canopy mortality as a result of increasing temperature [@senfCanopyMortalityHas2018a] has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019, @defrenneMicroclimateModeratesPlant2013]

@@

Understanding thermal sensitivity along the vertical canopy profile, from the understory to the top of the canopy will provide crucial insights for how raising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions [with additional feedbacks to forest structure and microclimates].

>>>>

Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF). 
There is also increasing severity and frequency of heat waves (@perkinsIncreasingFrequencyIntensity2012).
This is expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will have a critical influence on the future of trajectory of climate change [@bonan_ecological_2016, Bonan 2008].
These changes, alongside climate change-driven disturbances such as drought, deforestation and fire, may contribute to increasing tree mortality in forests around the world [@mcdowellMechanismsPlantSurvival2008, @allenUnderestimationGlobalVulnerability2015, Brienen et al. 2015].

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**
Larger trees form canopies that maintain vertically-stratified forest ecosystems, and play a crucial role in regulating macro-microclimate. (Nakamura et al. 2017, Ozanne et al. 2003)  
With increase in temperatures globally, we're seeing that larger trees suffer more during drought [@bennettLargerTreesSuffer2015, @stovallTreeHeightExplains2019] 
Increasing mortality of large trees, sometimes coupled with reduced tree growth rates, is driving a global trend towards more disturbed forests with younger trees (McDowell et al.2020).

**The structure and microclimates of global forests are changing due to unprecedented disturbance rates.**
Temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018, @huangAirTemperatureOptima2019].Additionally, loss of canopy cover as a result of increasing temperature and related forest disturbances [@senfCanopyMortalityHas2018a] is associated with reductions in canopy structural complexity, and altered microclimates and surface energy balance components (Zellweger et al. 2020; Stark et al. 2020 ), this has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019, @defrenneMicroclimateModeratesPlant2013]

(Marielle--"Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions (Stark et al. 2020)".)

@@

With shifting forest dynamics, it is becoming more important than ever to understand thermal sensitivity along the vertical canopy profile, from the understory to the top of the canopy, this will provide crucial insights for how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions [with additional feedbacks to forest structure and microclimates].
